---
  title: "Reproducible Research: Peer Assessment 1"
  output: html_document
  keep_md: true
---

## Loading and preprocessing the data

####Loading the needed libraries

```{r}
library(ggplot2)
library(plyr)
library(Hmisc)
library(lattice)
```

####Loading and preprocessing the data

```{r}
activity <- read.csv("r/activity.csv")
head(activity)

activity$day <- weekdays(as.Date(activity$date))
activity$date<- as.POSIXct(activity$date, format="%Y-%m-%d")
head(activity)
str(activity)
```

## What is mean total number of steps taken per day?

#### Total number of steps taken per day

```{r}
stepsPerDay <- tapply(activity$steps, activity$date, sum)


head(stepsPerDay)
```

#### Histogram of the total number of steps taken each day

```{r}
hist(stepsPerDay)
```

```{r}
stepsPerDayMean <- mean(stepsPerDay, na.rm = TRUE)
stepsPerDayMedian <- median(stepsPerDay, na.rm = TRUE)
```

* Mean steps per Day: `r stepsPerDayMean`
* Median steps per Day:  `r stepsPerDayMedian`



## What is the average daily activity pattern?

#### removing the NAs

```{r}
clean <- activity[!is.na(activity$steps),]
```

#### Create average number of steps per interval
```{r}
intervalTable <- ddply(clean, ~ interval, plyr::summarize, Avg = mean(steps))
```

#### Create line plot of average number of steps per interval
```{r}
ggplot(intervalTable, aes(x=interval, y=Avg)) +
  geom_line()+xlab("Interval") +
  ylab("Average Number of Steps") + 
  ggtitle("Average Number of Steps per Interval")
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

####Maximum steps by interval
```{r}
maxSteps <- max(intervalTable$Avg)
```

####Which interval contains the maximum average number of steps
```{r}
intervalTable[intervalTable$Avg==maxSteps,1]
```

The 5-minute interval which had the maximum number of steps was the 
`r intervalTable[intervalTable$Avg==maxSteps,1]` interval.


#### Total number of missing values in the dataset : `r nrow(activity[is.na(activity$steps),])`
```{r}
nrow(activity[is.na(activity$steps),])
```

## Imputing missing values

####Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r}
activityImputed <- activity
activityImputed$steps <- impute(activity$steps)  
```
#### I use the median value across the entire dataset as the imputed value (default)
#### I did not mean as the imputed value as imputing with mean will result in no change in the mean value before and after imputing 

#### CHeck that number of NAs = 0 now
```{r}
nrow(activityImputed[is.na(activityImputed$steps),])
```

#### Create histogram
```{r}
stepsPerDayImputed <- tapply(activityImputed$steps, activityImputed$date, sum)

hist(stepsPerDayImputed)
head(stepsPerDayImputed)
```

#### Calculate and report the mean and median total number of steps taken per day

```{r}
stepsPerDayMeanImputed <- mean(stepsPerDayImputed)
stepsPerDayMedianImputed <- median(stepsPerDayImputed)
```

* Mean steps per Day (imputed): `r stepsPerDayMeanImputed`
* Median steps per Day (imputed):  `r stepsPerDayMedianImputed`

#### What is the impact of imputing missing data on the estimates of the total daily number of steps?
#### Ans : The mean and median steps after imputation is lower compared to the respective values without imputation


## Are there differences in activity patterns between weekdays and weekends?

#### Create new category based on the days of the week
```{r}
activityImputed$DayCategory <- ifelse(activityImputed$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```

```{r}
intervalTable2 <- aggregate(steps ~ interval + DayCategory, data=activityImputed, mean)

ggplot(intervalTable2, aes(interval, steps)) + 
  geom_line() + 
  facet_grid(DayCategory ~ .) +
  xlab("5-minute interval") + 
  ylab("avarage number of steps")
```

#### Yes, there are differences

####the activity is more uniform / consistent across time intervals during weekends
####For weekdays, there is a larger range of steps. A possible explanation is working over the weekdays and there are defined periods on the job where higher activity is expected